---
title: "Assignment 2"
author: "Adam Weiner"
output: html_document
---
##Setup
```{r}
library(caret)
library(class)
library(dplyr)
```

```{r}
bank <- read.csv("/Users/adamweiner/Downloads/UniversalBank-1.csv")

str(bank)

summary(bank)
```

##Question 1: Partition data 60/40, predictors, and conversion

##Data Prep
```{r}
# Remove ID and ZIP
bank2 <- bank[, !(names(bank) %in% c("ID", "ZIP.Code"))]

# Convert ED
bank2$Education <- as.factor(bank2$Education)

#Dummy Vars
dummies <- dummyVars(Personal.Loan ~ ., data = bank2)

bank_dummy <- data.frame(
  predict(dummies, newdata = bank2)
)

#Response Variable
bank_dummy$Personal.Loan <- as.factor(bank2$Personal.Loan)

str(bank_dummy)

```
### Training and Validation
```{r}
set.seed(123)

train_index <- createDataPartition(
  bank_dummy$Personal.Loan,
  p = 0.60,
  list = FALSE
)

train_data <- bank_dummy[train_index, ]
validation_data <- bank_dummy[-train_index, ]

dim(train_data)
dim(validation_data)

```

### Separate predictors and response
```{r}
train_x <- train_data[, !(names(train_data) %in% c("Personal.Loan"))]
validation_x <- validation_data[, !(names(validation_data) %in% c("Personal.Loan"))]

train_y <- train_data$Personal.Loan
validation_y <- validation_data$Personal.Loan
```

###Consider this Customer
```{r}
new_customer <- data.frame(
  Age = 40,
  Experience = 10,
  Income = 84,
  Family = 2,
  CCAvg = 2,
  Education.1 = 0,
  Education.2 = 1,
  Education.3 = 0,
  Mortgage = 0,
  Securities.Account = 0,
  CD.Account = 0,
  Online = 1,
  CreditCard = 1
)

# k-NN classification where k = 1
knn_pred_k1 <- knn(
  train=train_x,
  test = new_customer,
  cl = train_y,
  k = 1
)

knn_pred_k1
```
### Question 1 Write-up
Using the k-NN classification where k=1 with all predictors except for ID and ZIP Code, the customer was found to be classified as 0. This indicates that the model predicted the customer would not accept the personal loan offer based off their qualifications. 


### Question 2: Choice of K that balances overfitting and ignores predictor info

```{r}
# Testing values of k
k_values <- seq(1,25, by = 2)

accuracy <- c()

for(k in k_values){
  pred <- knn(
    train = train_x,
    test = validation_x,
    cl = train_y,
    k = k
  )
  accuracy <- c(
    accuracy,
    mean(pred == validation_y)
  )
}

#Results Table
k_results <- data.frame(
  k = k_values,
  Accuracy = accuracy
)

k_results
```

### Best k
```{r}
best_k <- k_results$k[which.max(k_results$Accuracy)]

best_k
```
### Question 2 Write-up
Multiple odd values of k were tested in the validation data. I selected the highest validation accuracy that was when k = 3 with a value of 0.9070. So, k = 3 was selected as it has the best balance between overfitting and the loss of predictor information that comes with larger values of k.


### Question 3 Confusion Matrix
```{r}
pred_best <- knn(
  train = train_x,
  test = validation_x,
  cl = train_y,
  k = best_k
)

confusionMatrix(
  pred_best,
  validation_y,
  positive = "1"
)
```

### Question 3 Write-up
In using the best value of k = 3, a confusion matrix was created for the validation data. The model identified 1750 customers who did not accept the loan offer and 64 customers who did accept it correctly. The model incorrectly identified 128 customers who accepted the loan when they shouldn't have and 58 who did not accept when they should have. The overall accuracy of the classification in the model was 0.907, which means the model correctly classified 90.70% of the customers in the validation data. 


###Question 4 Customer with Best k
```{r}
knn_pred_best_customer <- knn(
  train = train_x,
  test = new_customer,
  cl = train_y,
  k = best_k
)

knn_pred_best_customer
```

### Question 4 Write-up
I used the best value of k=3 and confirmed that the customer was classified as 0, which indicated that the model predicts the customer would not accept the personal loan offer here. Changing from k=1 to k=3 didn't change the prediction for this specific customer. 

###Question 5 

###Repartition Data
```{r}
set.seed(123)

#Split 1 50%
train_index_50 <- createDataPartition(
  bank_dummy$Personal.Loan,
  p = 0.50,
  list = FALSE
)

train_50 <- bank_dummy[train_index_50, ]
remaining_50 <- bank_dummy[-train_index_50, ]

#Split 2 60% of remaining 50%
validation_index_30 <- createDataPartition(
  remaining_50$Personal.Loan,
  p = 0.60,
  list = FALSE
)

validation_30 <- remaining_50[validation_index_30, ]
test_20 <- remaining_50[-validation_index_30, ]

dim(train_50)
dim(validation_30)
dim(test_20)
```

###Predictors and Response Variables
```{r}
#Training predictors and responses
train50_x <- train_50[, !(names(train_50) %in% c("Personal.Loan"))]
train50_y <- train_50$Personal.Loan

#Validation
validation30_x <- validation_30[, !(names(validation_30) %in% c("Personal.Loan"))]
validation30_y <- validation_30$Personal.Loan

#Test
test20_x <- test_20[, !(names(test_20) %in% c("Personal.Loan"))]
test20_y <- test_20$Personal.Loan
```

#Predictions with k=3
```{r}
train_pred <- knn(
  train = train50_x,
  test = train50_x,
  cl = train50_y,
  k = best_k
)

validation_pred <- knn(
  train = train50_x,
  test = validation30_x,
  cl = train50_y,
  k = best_k
)

test_pred <- knn(
  train = train50_x,
  test = test20_x, 
  cl = train50_y,
  k = best_k
)
```

###Confusion Matrix
```{r}
#Training
confusionMatrix(
  train_pred,
  train50_y,
  positive = "1"
)

#Validation
confusionMatrix(
  validation_pred,
  validation30_y,
  positive = "1"
)

#Test
confusionMatrix(
  test_pred,
  test20_y,
  positive = "1"
)
```

### Question 5 Write-up
The dataset was repartitioned into training, validation, and test sets (50%, 30%, 20%). I then used the best k = 3 and created confusion matricies for each of the three sets. The training dataset had produced the highest level of accuracy at 94.68%, and the validation and test sets produced 91.12% and 91.30% respectively. The training dataset had produced the highest level of accuracy since the model itself was built off of these exact observations. The validation and test sets were smaller than that of the training, but by only roughly 3.5%, and were very similar in their accuracy levels. This may suggest that the model does well at generalizing the unseen data points and does not overfit the data either. Having small differences between each of the datasets is as expected since each set contains different observations. 
