---
title: "Assignment 1"
output: html_document
---

## Load Data
The dataset I used contains NBA player statistics for the 2025-2026 season. I obtained it from Basketball Reference. It contains both categorical and quantitative variables. 
```{r}
library(readr)

nba <- read_csv("Downloads/Untitled spreadsheet - Sheet1.csv")
```

## View data
I used the head function to view the first six observations in the data to be sure the data was imported correctly. It also showed me an overview of the variables I would be using in my analysis.
```{r}
head(nba)
```

## Descriptive Statistics (Quantitative Variables)
The summary statistics provided are of three quantitative variables. I chose to do points per game (PTS), assists per game (AST), and total rebounds per game (TRB). The summaries of these variables provide information on the minimum, first quarter, median, mean, third quarter, and maximum values. 
```{r}
summary(nba$PTS)
summary(nba$AST)
summary(nba$TRB)
```
## Descriptive Statistics (Categorical Variables)
The two frequency tables provide information on the two categorical variables of player position and team. The two tables show the distribution of players across the 5 positions in basketball and the 30 teams. 
```{r}
table(nba$Pos)
table(nba$Team)
```
## Variable Transformation
Assist to turnover ration (AST_TOV) variable made by dividing assists per game by turnovers per game, measures players efficiency in creating assists while limiting turnovers. 
```{r}
nba$AST_TOV <- nba$AST / nba$TOV
head(nba$AST_TOV)
```
## Histogram
This histogram shows the distribution of points per game for all NBA players. Most of the players in the league scored between 0 and 10 points, whereas a very small percentage scored above 25 points per game. This is as expected. 
```{r}
hist(nba$PTS,
     main = "Distribution of Points Per Game",
     xlab = "Points Per Game")
```

## Scatterplot
The scatterplot shows the relationship between assists per game and points per game. The plot suggests a positive relationship with players who average a higher number of assists also average more points per game. However, most players average a low number of assists and points per game still. 
```{r}
plot (nba$AST,
      nba$PTS,
      main = "Assists Per Game vs Points Per Game",
      xlab = "Assists Per Game",
      ylab = "Points Per Game")
```